---
title: "Offline GBC PheWAS mediation follow-up"
author: "Gibran Hemani"
date: "2026-10-09"
execute:
  cache: false
  freeze: false
  echo: false
  warning: false
  message: false
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
root <- if (file.exists("data/offline-g2-inputs.rds")) "." else ".."
knitr::opts_knit$set(root.dir = normalizePath(root))
```

```{r}
source("scripts/R/offline-pipeline.R")
input <- Sys.getenv("GBC_G2_INPUT", "data/gbc_phewas_results-g2.Rdata")
analysis <- run_offline_analysis(input)
comparison <- analysis$comparisons
pooled <- analysis$pooled
decomposition <- analysis$decomposition
interval <- function(b, lo, hi) ifelse(is.finite(b), sprintf("%.3g [%.3g, %.3g]", b, lo, hi), "Unavailable")
show_table <- function(x) knitr::kable(x, row.names = FALSE, digits = 4)
```

## Inventory and local inputs

The report rebuilds from saved associations, with remote extraction, downloads and
package installation disabled. Old regional `exp_sets`, reduced `outdat`, old MR
results and knitr result caches are not inputs to the corrected analysis.

There are **`r nrow(comparison)` unique pairs**, **`r nrow(pooled)` mediators** and
**`r length(unique(comparison$target_snp))` retained target SNPs**. The saved follow-up
had `r analysis$inputs$inventory_counts$n_saved_pair_rows` rows; repeated rows do not
add comparisons. The original PheWAS covered eight lead SNPs. The two without retained
saved follow-up associations are accounted for below; the saved data do not establish
why they were not retained.

```{r}
show_table(analysis$tables$lead_inventory)
```

These GBC lead variants are not generally weak associations: seven of eight have
P < 5e-8; rs2131242 is below that significance threshold (saved outcome P =
`r format_probability(analysis$tables$lead_inventory$gbc_pval[analysis$tables$lead_inventory$target_snp == "rs2131242"])`).
The original report quoted approximately 4.81e-7; the saved outcome association
used here differs, and both values are below genome-wide significance.
Genome-wide significance alone does not establish instrument validity.

## Exact-hit comparisons

Every target estimate uses exactly its named SNP in `int_chd`, harmonised with the
full saved GBC outcome `outgwasf`. The comparator is rebuilt from that mediator's
saved instruments, excluding the target and **±1 Mb**, inclusive, before outcome
matching and harmonisation. Saved coordinate mismatches and missing instrument
coordinates are excluded. A geographic window does not prove LD independence;
there is no fetched LD reference or genome-build conversion. The supplement includes
±2 Mb and additional chromosome 6 25–34 Mb MHC exclusions where relevant.

TwoSampleMR action 3 removes every A/T and C/G palindrome regardless of frequency.
Non-palindromic strand complements and allele reversals are harmonised; incompatible
alleles, unvalidated proxies, conflicting duplicates and invalid SEs yield explicit
unavailability. The recorded target effect allele orients both target associations.
A usable total GBC association may remain when a mediator component is unusable.

The exact target uses a Wald ratio. Comparators use a Wald ratio for one instrument
or IVW with multiplicative random effects for multiple instruments. Conventional
first-order ratio SEs treat exposure effects as fixed; beta/SE-derived F statistics
are available in the CSV, and measure strength rather than validity or independence.
Decomposition below propagates uncertainty in both multiplicands. Specific mediator
transformations/units were not preserved: effects use **saved GWAS units, specific
unit unavailable**, with target and comparator matched by mediator GWAS ID.
Comparison panels use their own scales; magnitudes across mediators are not comparable.

```{r}
show_table(data.frame(SNP = comparison$target_snp, mediator = comparison$mediator,
  EA = comparison$effect_allele,
  exact_CI = interval(comparison$target_b, comparison$target_lo, comparison$target_hi),
  other_CI = interval(comparison$comparator_b, comparison$comparator_lo, comparison$comparator_hi),
  n = comparison$comparator_nsnp, Q_P = format_probability(comparison$q_pval),
  Q_BH_P = format_probability(comparison$q_padj), interpretation = comparison$interpretation))
```

```{r results='asis'}
for (path in analysis$plots$comparisons) cat(sprintf("\n![Exact-hit comparison](../%s)\n\n", path))
```

## Pooled mediator MR

Pooled MR excludes the **union of every target locus for each mediator directly
from its saved instruments**. A SNP excluded for one target cannot return through
another target's complement. This distinct summary is never substituted for the
per-target comparator used in decomposition.

```{r}
show_table(data.frame(mediator = pooled$mediator, targets = pooled$n_targets,
  excluded_before_matching = pooled$n_excluded_target + pooled$n_excluded_region,
  excluded_after_matching = pooled$n_excluded_after_outcome_matching,
  n = pooled$nsnp, method = pooled$method,
  estimate_CI = interval(pooled$b, pooled$lo, pooled$hi), P = format_probability(pooled$pval),
  BH_P = format_probability(pooled$pval_bh), minimum_F = pooled$F_min, reason = pooled$reason))
```

```{r results='asis'}
platelet <- pooled[grepl("platelet count", pooled$mediator, ignore.case = TRUE), ]
for (k in seq_len(nrow(platelet))) cat(sprintf(
  "Platelet-count candidate %s: n=%s, estimate %s, nominal P=%s and pooled-family BH P=%s.\n\n",
  platelet$mediator_id[k], platelet$nsnp[k], interval(platelet$b[k], platelet$lo[k], platelet$hi[k]),
  format_probability(platelet$pval[k]), format_probability(platelet$pval_bh[k])))
for (path in analysis$plots$pooled) cat(sprintf("\n![Pooled mediator MR](../%s)\n\n", path))
```

## Total, indirect and residual direct components

For each pair: **t = G→GBC**, **i = a × b**, **d = t − i**, where a is the exact
G→mediator association and b is that target's other-instruments MR. Components share
GBC log-odds units per recorded target effect allele. Opposing signs are retained;
missing components are never zero. The pipeline validates t = i + d for every
complete pair.

Approximate normal 95% CIs use a first-order delta method with zero covariances:
Var(i) = b²Var(a) + a²Var(b), Var(d) = Var(t) + Var(i). General formulas add
2abCov(a,b) to Var(i), and subtract 2bCov(a,t) + 2aCov(b,t) from Var(d). Saved
summary data do not identify these covariances. The supplement uses hypothetical
equicorrelations −0.3, 0 and +0.3 among a, b and t; these are positive-definite
matrices, not estimates of empirical covariance or sample overlap.

These are **model-based approximate components**, not identified causal natural
effects. Binary traits, non-collapsibility, nonlinearity, population differences,
LD and instrument validity limit causal interpretation. Indirect effects for
separate correlated mediators must not be summed. No mediation percentage is reported.

```{r}
show_table(data.frame(SNP = decomposition$target_snp, mediator = decomposition$mediator,
  EA = decomposition$effect_allele,
  total_CI = interval(decomposition$t, decomposition$t_lo, decomposition$t_hi),
  indirect_CI = interval(decomposition$i, decomposition$i_lo, decomposition$i_hi),
  residual_CI = interval(decomposition$d, decomposition$d_lo, decomposition$d_hi),
  indirect_P = format_probability(decomposition$i_pval), indirect_BH_P = format_probability(decomposition$i_padj),
  residual_P = format_probability(decomposition$d_pval), residual_BH_P = format_probability(decomposition$d_padj),
  reason = decomposition$decomposition_reason))
```

```{r results='asis'}
for (path in analysis$plots$decomposition) cat(sprintf("\n![Effect decomposition](../%s)\n\n", path))
```

## Interpretation and testing families

Q indicates incompatibility with the common-effect/full-mediation model under its
assumptions. Invalid target/comparator instruments, residual LD or population/scale
differences may explain incompatibility; Q does not uniquely locate a direct pathway.
A non-significant comparison means **no detected heterogeneity; mediation unresolved**.
It proves neither mediation nor equivalence. Comparator precision and instrument
strength are recorded in the CSV.

There are `r sum(comparison$q_padj < 0.05, na.rm=TRUE)` BH-incompatible pairs among
`r sum(is.finite(comparison$q_pval))` tested unique pairs in the retained inventory of
`r nrow(comparison)` pairs. Each named family separately applies BH at 0.05 to its
available tests; this does not claim global control across separate families.
Indirect/residual tests use approximate normal delta-method intervals. The two window/MHC sensitivity scenarios each use their own separately named
33-pair BH family; covariance scenarios describe approximate interval sensitivity.
Per-SNP counts describe pair tests and are not additional SNP-level hypothesis tests.

```{r}
show_table(analysis$summary$families)
show_table(analysis$summary$per_snp)
```

Every row is traceable by mediator ID and SNP to the saved exact-association export,
retained comparator SNP list, exclusion counts and explicit missingness reasons.
All empirical pairs happen to be usable under the primary policy.
The unavailable example below is deliberately labelled as an illustrative
missing-comparator scenario, excluded from empirical counts.

```{r}
z <- analysis$tables$representative_relationships
show_table(z[c("example", "source_kind", "mediator_id", "target_snp", "interpretation", "precision", "reason")])
```
 Synthetic tests
validate unavailable inputs, allele reversals, complete mediation, residual effects,
opposing pathways, increasing uncertainty and covariance sensitivities.

## Downloads and reproduction

- [Comparison PDF](../figures/phewas_followup_results-g2.pdf) / [first PNG page](../figures/phewas_followup_results-g2.png)
- [Decomposition PDF](../figures/phewas_decomposition-g2.pdf) / [first PNG page](../figures/phewas_decomposition-g2.png)
- [Pooled MR PDF](../figures/phewas_pooled_mr-g2.pdf)
- [Comparison CSV](../results/g2/comparisons.csv), [pooled MR CSV](../results/g2/pooled_mr.csv), [decomposition CSV](../results/g2/decomposition.csv)
- [Representative traceable relationships](../results/g2/representative_relationships.csv)
- [Legacy pooled diagnostics](../results/g2/pooled_legacy_diagnostic.csv), [complement-union bug counts](../results/g2/pooled-complement-union-diagnostic.csv)
- [Window/MHC sensitivity](../results/g2/window_sensitivity.csv), [covariance sensitivity](../results/g2/covariance_sensitivity.csv)
- [Original-hit inventory](../results/g2/lead_inventory.csv), [saved exact associations](../results/g2/exact_saved_associations.csv)
- [Testing families](../results/g2/inference_families.csv), [per-SNP summary](../results/g2/per_snp_summary.csv)
- [Provenance](../results/g2/provenance.csv), [package versions](../results/g2/package_versions.csv), [session information](../results/g2/session-info.txt)

From the repository root, using the existing local `data/gbc_phewas_results-g2.Rdata`:

```bash
Rscript scripts/run-offline-g2.R
quarto render scripts/phewas-followup-g2.qmd --no-cache
Rscript tests/run-tests.R
```

A compact lossless subset of the required saved associations is included in
`data/offline-g2-inputs.rds`. The source checksums and extraction counts are in
[Snapshot provenance](../results/g2/input-snapshot-provenance.csv). No new empirical
data were acquired. For a checkout without the original Rdata:

```bash
Rscript scripts/run-offline-g2.R data/offline-g2-inputs.rds
GBC_G2_INPUT=data/offline-g2-inputs.rds quarto render scripts/phewas-followup-g2.qmd --no-cache
```

No remote installation or data access is needed. The older G1 workflow is unchanged.
Each render recomputes all tables/figures and bypasses old caches.

```{r}
show_table(analysis$provenance)
show_table(analysis$versions)
```

<style>
.cell-output-display { overflow-x: auto; }
table { font-size: 0.82rem; }
figure img { max-width: 100%; height: auto; }
</style>
